Partitioned Local Depth (PaLD) Clustering Analyses in R

### 2022-03-13 r-journal/manuscript.Rmd
update paper

* Describe PaLD (Ken & Kate?)
* Cite PaLD

@@

1. Helper functions to organize data into the correct format, into distance matrices and then contribution matrices
2. Functions that convert a contribution matrix into a variety of useful formats, including partitioned local depths, clusters, and graphs
3. Plotting functions

@@

While it is not a necessity, the \CRANpkg{pald} package is designed to function well with the pipe operator, `%>%`, from the \citet{magrittr} \CRANpkg{magrittr} package in that the first argument of each function is the data. This functionality will be demonstrated below.

@@

For demonstration purposes, below is a sample data frame with two variables, `x1` and `x2`. The methods put forth here work on data frames with higher dimensions, as described in the **Examples** section, we are simply choosing a small data frame here for demonstration purposes.

```{r}
library(pald)
d <- data.frame(
  x1 = c(6, 8, 11, 16, 4),
  x2 = c(5, 4, 13, 7, 18)
)
rownames(d) <- c("A", "B", "C", "D", "E")
```

The first step needed to calculate the partitioned local depths is to construct a _distance matrix_. If the data are already in this form, the user can skip to the next step. The `get_distance_matrix()` function converts an input data frame into a distance matrix, as demonstrated below.

```{r}
get_distance_matrix(d)
```
This will create an $n\times n$ distance matrix, where $n$ corresponds to the number of observations in the original data frame, in this example $n = 5$. By default, the distance matrix is _scaled_, is possible to create a distance matrix that is not scaled by changing the `scale` argument to `FALSE`.

```{r}
get_distance_matrix(d, scale = FALSE)
```

 As mentioned previously, the functions in this package are designed to work with the \CRANpkg{magrittr} pipe operator, `%>%`, so the same code as above could be written utilizing this format.

```{r}
d %>%
  get_distance_matrix()
```

This distance matrix can then be passed to the `get_contribution_matrix()` function in order to calculate a matrix of contributions. Again, if the user begins with a distance matrix, they can skip the first step and simple input the contribution matrix into this function.

```{r}
distance_matrix <- get_distance_matrix(d)
get_contribution_matrix(distance_matrix)
```

Again, the \CRANpkg{magrittr} pipe can be used as follows.

```{r}
d %>%
  get_distance_matrix() %>%
  get_contribution_matrix()
```

The _contribution matrix_ output by the `get_contribution_matrix()` is the main input for the majority of the remaining functions.

### Functions that convert a contribution matrix into useful formats

From the _contribution matrix_, a variety of useful quantities can be calculated. Below, we create a contribution matrix using the functions described in the previous section.

```{r}
d %>%
  get_distance_matrix() %>%
  get_contribution_matrix() -> contribution_matrix
```

To calculate the _clusters_ that each point will fall into, we can use the `get_clusters()` function. This will output a data frame with two columns, the first will correspond to the `point`, as identified by the row name of the original input data frame, `d`, the second will identify the `cluster` that each point belongs to.

```{r}
get_clusters(contribution_matrix)
```

@@

The `get_depths()` function calculates the _depths_ of each point, outputting a data frame with two columns, `point` indicating the point, as identified by the row name of the original data frame, `d`, and `depth` indicating the depth of the point. The data frame is arranged by depth, with the deepest point listed first.

```{r}
get_depths(contribution_matrix)
```

@@

The `get_bound()` function will calculate the _bound_ of the contribution matrix.

```{r}
get_bound(contribution_matrix)
```

In this case, the bound is 0.14.

@@

```{r}
any_isolated(contribution_matrix)
```

In this case, there are no isolated points.

The `get_pald_cluster_matrix()` will calculate a matrix of partitioned local depth clusters. This function contains an argument `keep_all_edges` that indicates whether all edges should be kept. The default value is `FALSE`, indicating that edges that are less than the expectation will be dropped, allowing a clear clustering of points. Changing this argument to `TRUE` results in the pairwise minimum of the contribution matrix and the transpose of the contribution matrix.

```{r}
get_pald_cluster_matrix(contribution_matrix, keep_all_edges = TRUE)
```

Leaving the `keep_all_edges` as the default, `FALSE` will take this pairwise minimum seen above and drop the edges that are less than the expectation.

```{r}
get_pald_cluster_matrix(contribution_matrix)
```

Finally, the `get_pald_graph()` function takes the contribution matrix and creates an \CRANpkg{igraph} object, a graph that describes the relationship between the points.

```{r}
get_pald_graph(contribution_matrix)
```

@@

The final category of function is functions for data visualization. We can begin by visualizing the points in data frame `d` (Figure \ref{fig:fig1}).

```{r fig1, fig.cap = "Visualize the points from data frame `d`", message = FALSE, warning = FALSE}
library(ggplot2)
ggplot(d, aes(x1, x2)) +
  geom_text(label = rownames(d))
```

We can then pass the contribution matrix to the `plot_pald_graph()` function to view the relationship between points (Figure \ref{fig:fig2}). By default, this will invoke the `layout_nicely()` function from \CRANpkg{igraph} to determine the layout of the graph.

```{r fig2, fig.cap = "PaLD graph displaying the relationship between the points in data frame `d`"}
d %>%
  get_distance_matrix() %>%
  get_contribution_matrix() %>%
  plot_pald_graph()
```

The `layout` argument allows the user to pass a matrix to dictate the layout of the graph. For example, if we wanted the graph to match the visualization displayed in Figure \ref{fig:fig1}, we can pass `as.matrix(d)`, or a matrix of the data frame `d` to the `layout` argument  (Figure \ref{fig:fig3}.  
```{r fig3, fig.cap = "PaLD graph displaying the relationship between the points in data frame `d`, matching the original layout in Figure 1"}
d %>%
  get_distance_matrix() %>%
  get_contribution_matrix() %>%
  plot_pald_graph(layout = as.matrix(d))
```

This `plot_pald_graph()` function will also permit parameters that can be passed to the `plot.igraph()` function. For example, to add axes to the graph, the user can pass the `axes = TRUE` argument to the `...` in the `plot_pald_graph()` function (Figure \ref{fig:fig4}).

```{r fig4, fig.cap = "PaLD graph displaying the relationship between the points in data frame `d`, matching the original layout in Figure 1, adding axes"}
d %>%
  get_distance_matrix() %>%
  get_contribution_matrix() %>%
  plot_pald_graph(layout = as.matrix(d),
                  axes = TRUE)
```

@@

The first example data frame is a subset of tissue gene expression data from \citet{zilliox2007gene}, \citet{mccall2011gene}, and \citet{mccall2014gene}, obtained from the **tissuesGeneExpression** bioconductor package \citep{tissue}. This data set is included in the \CRANpkg{pald} package in an object called `tissue`.

The `tissue` object is a matrix with 189 rows, each with a corresponding tissue, such as `colon`, `kidney` or `cerebellum`. There are 22,215 columns corresponding to gene expression data from each of these rows.

We will first calculate the contribution matrix for this data frame.

```{r}
tissue_contribution_matrix <- tissue %>%
  get_distance_matrix(scale = FALSE) %>%
  get_contribution_matrix()
```

We can then use this contribution matrix to display the relationship between tissue samples using the `plot_pald_graph()` function (Figure \ref{fig:fig5}).

```{r fig5, fig.cap = "PaLD clustering of tissue data"}
plot_pald_graph(tissue_contribution_matrix)
```

The `get_clusters()` function can be used to identify the clusters of each tissue sample. Since the output is a data frame, we can summarize the clusters using commonly used data analysis techniques. For demonstration purposes, we will use the \CRANpkg{dplyr} package to summarize the contribution of clusters.

```{r}
library(dplyr)
get_clusters(tissue_contribution_matrix) %>%
  group_by(cluster, point) %>%
  count()
```

@@

The \CRANpkg{pald} includes two data frames generated from the scikit-learn Python package \citep{pedregosa2011scikit}, `noisy_moons` and `noisy_circles`.

The `noisy_moons` data frame consists of 500 rows and two columns.

```{r}
moons_contribution_matrix <- noisy_moons %>%
  get_distance_matrix() %>%
  get_contribution_matrix()
```

When plotting the `noisy_moons` PaLD graph, we want the layout to match the layout of the original data, so we will pass `as.matrix(noisy_moons)` to the `layout` parameter in the `plot_pald_graph()` function. Additionally, here the row names are meaningless, they just correspond to the location of the generated data, so we can remove the labels on the plot by passing `vertex.label = NA` to the `...` of the `plot_pald_graph()` function.

```{r, eval = FALSE}
plot_pald_graph(moons_contribution_matrix,
                layout = as.matrix(noisy_moons),
                vertex.label = NA)
```

The `noisy_circles` data frame consists of 500 rows and two columns. We can create the contribution matrix for this data frame using the same methods as used for the `noisy_moons` data.

```{r}
circles_contribution_matrix <- noisy_circles %>%
  get_distance_matrix() %>%
  get_contribution_matrix()
```

Similarly, we will pass the layout and remove vertex labels for the noisy circles PaLD plot.

```{r, eval = FALSE}
plot_pald_graph(circles_contribution_matrix,
                layout = as.matrix(noisy_circles),
                vertex.label = NA)
```

Because these are \CRANpkg{igraph} objects, they can be combined as you would normally combine a plot. For example, we could add the line `par(mfrow = c(1, 2))` above the plot calls to create an output where both plots are displayed side by side (Figure \ref{fig:fig6}).

```{r fig6, fig.cap = "PaLD clustering of noisy moons (left) and noisy circles (right) data"}
par(mfrow = c(1, 2))
plot_pald_graph(moons_contribution_matrix,
                layout = as.matrix(noisy_moons),
                vertex.label = NA)
plot_pald_graph(circles_contribution_matrix,
                layout = as.matrix(noisy_circles),
                vertex.label = NA)
```

>>>>

* Describe PaLD (Ken & Kate?) [@berenhaut2022social]

@@

1. Helper functions to organize data into the correct format, into distance matrices and then cohesion matrices
2. Functions that convert a cohesion matrix into a variety of useful formats, including partitioned local depths, clusters, and graphs
3. Plotting functions

@@

While it is not a necessity, the \CRANpkg{pald} package is designed to function well with the pipe operator, `|>`. This functionality will be demonstrated below.

@@

For demonstration purposes, below is a sample data frame with two variables, `x1` and `x2`. The methods put forth here work on data frames with higher dimensions, as described in the **Examples** section; we are simply choosing a small data frame here for demonstration purposes.

```{r}
library(pald)
df <- data.frame(
  x1 = c(6, 8, 11, 16, 4),
  x2 = c(5, 4, 13, 7, 18)
)
rownames(df) <- c("A", "B", "C", "D", "E")
```

The first step needed to calculate the partitioned local depths is to construct a _distance matrix_. If the data are already in this form, the user can skip to the next step. The `dist()` function converts an input data frame into a distance matrix, as demonstrated below.

```{r}
d <- dist(df)
```

This will create an $n\times n$ distance matrix, where $n$ corresponds to the number of observations in the original data frame, in this example $n = 5$. 
<!-- TODO: is this nxn? looks like (n-1)x(n-1) -->
This distance matrix can then be passed to the `cohesion_matrix()` function in order to calculate the pairwise cohesion values. Cohesion is an interpretable probability that reflects the strength of alignment of two points. Again, if the user begins with a distance matrix, they can skip the first step and simply input the distance matrix into this function.

```{r}
d <- dist(df)
cohesion_matrix(d)
```

Equivalently, the user can use the native pipe `|>` as follows.

```{r}
df |>
  dist() |>
  cohesion_matrix()
```

The _cohesion matrix_ output by the `cohesion_matrix()` is the main input for the majority of the remaining functions.

### Functions that convert a cohesion matrix into useful formats

From the _cohesion matrix_, a variety of useful quantities can be calculated. Below, we create a cohesion matrix using the functions described in the previous section.

```{r}
d |>
  dist() |>
  cohesion_matrix() -> cohesion
```

To calculate the _clusters_ that each point will fall into, we can use the `community_clusters()` function. This will output a data frame with two columns, the first will correspond to the `point`, as identified by the row name of the original input data frame, `df`, the second will identify the `cluster` that each point belongs to.

```{r}
community_clusters(cohesion)
```

@@

The `local_depths()` function calculates the _depths_ of each point, outputting a vector of local depths. Local depth is an interpretable probability which reflects aspects of relative position and centrality via distance comparisons (i.e., $d(z, x) < d(z, y)$).

```{r}
local_depths(cohesion)
```

@@

The `strong_threshold()` function will calculate the cohesion threshold for strong ties. This is equal to half the average of the diagonal cohesion matrix. This is a threshold that may be used to distinguish between strong and weak ties.

```{r}
strong_threshold(cohesion)
```

In this case, the threshold is `0.14`.

@@

```{r}
any_isolated(cohesion)
```

Here, there are no isolated points.

The function `cohesion_strong()` will update the cohesion matrix to set all weak ties to zero (via the `strong_threshold()` function). Optionally, the matrix will also be symmetrized, with the default parameter `symmetric = TRUE`.

```{r}
cohesion_strong(cohesion)
```

Finally, the `community_graphs()` function takes the cohesion matrix and creates \CRANpkg{igraph} objects, graphs that describes the relationship between the points. This function will output a list of three objects:

* `G`: the weighted (community) graph whose edge weights are mutual cohesion
* `G_strong`: the weighted (community) graph consisting of edges for which mutual cohesion is greater than the threshold for strong ties
* `layout`: the graph layout, using the Fruchterman Reingold (FR) force-directed graph drawing for the graph `G`

```{r}
graphs <- community_graphs(cohesion)
graphs[["G_strong"]]
```

@@

The final category of function is functions for data visualization. We can begin by visualizing the points in data frame `df` (Figure \ref{fig:fig1}).

```{r fig1, fig.cap = "Visualize the points from data frame `df`", message = FALSE, warning = FALSE}
library(ggplot2)
ggplot(df, aes(x1, x2)) +
  geom_text(label = rownames(df))
```

We can then pass the cohesion matrix to the `plot_community_graphs()` function to view the relationship between points (Figure \ref{fig:fig2}).

```{r fig2, fig.cap = "PaLD graph displaying the relationship between the points in data frame `df`"}
df |>
  dist() |>
  cohesion_matrix() |>
  plot_community_graphs()
```

The `layout` argument allows the user to pass a matrix to dictate the layout of the graph. For example, if we wanted the graph to match the visualization displayed in Figure \ref{fig:fig1}, we can pass `as.matrix(df)`, or a matrix of the data frame `df` to the `layout` argument  (Figure \ref{fig:fig3}.  
```{r fig3, fig.cap = "PaLD graph displaying the relationship between the points in data frame `d`, matching the original layout in Figure 1"}
df |>
  dist() |>
  cohesion_matrix() |>
  plot_community_graphs(layout = as.matrix(df))
```

This `plot_community_graphs()` function will also permit parameters that can be passed to the `plot.igraph()` function. For example, to add axes to the graph, the user can pass the `axes = TRUE` argument to the `...` in the `plot_community_graphs()` function (Figure \ref{fig:fig4}).

```{r fig4, fig.cap = "PaLD graph displaying the relationship between the points in data frame `d`, matching the original layout in Figure 1, adding axes"}
df |>
  dist() |>
  cohesion_matrix() |>
  plot_community_graphs(layout = as.matrix(df),
                        axes = TRUE)
```

@@

The first example if from a subset of tissue gene expression data from \citet{zilliox2007gene}, \citet{mccall2011gene}, and \citet{mccall2014gene}, obtained from the **tissuesGeneExpression** bioconductor package \citep{tissue}. A cohesion matrix was created using this data set and is included the \CRANpkg{pald} package in an object called `tissue_cohesion_matrix`.

The `tissue_cohesion_matrix` object is a cohesion matrix with 189 rows and 189 columns.

We can use this contribution matrix to display the relationship between tissue samples using the `plot_community_graphs()` function (Figure \ref{fig:fig5}). For clarity of the display, we show how to only include a random set of the vertex labels. We can pass this random set through the `...` to the `plot.igraph` `vertex.label` parameter.

```{r fig5, fig.cap = "PaLD clustering of tissue data", fig.height = 9, fig.width = 7}
set.seed(1)

labels <- rownames(tissue_cohesion_matrix)
labels[sample(1:189, 125)] <- ""
plot_community_graphs(tissue_cohesion_matrix,
                      vertex.label = labels,
                      vertex.size = 4)
```

The `community_clusters()` function can be used to identify the clusters of each tissue sample. Since the output is a data frame, we can summarize the clusters using commonly used data analysis techniques. For demonstration purposes, we will use the \CRANpkg{dplyr} package to summarize the contribution of clusters.

```{r}
library(dplyr)
community_clusters(tissue_cohesion_matrix) %>%
  group_by(cluster, point) %>%
  count()
```

@@

The \CRANpkg{pald} includes three randomly generated data frames corresponding to plots from @berenhaut2022social:

* `exdata1` is a data set consisting of 8 points to recreate Figure 1 in @berenhaut2022social 
* `exdata2` is a data set consisting of 16 points to recreate Figure 2 in @berenhaut2022social 
* `exdata3` is a data set consisting of 240 points to recreate Figure 4D in @berenhaut2022social

Here, we will demonstrate how to use `exdata3`. These points were generated from bivariate normal distributions with varying means and variances.

```{r}
exdata_cohesion_matrix <- exdata3 |>
  dist() |>
  cohesion_matrix()
```

When plotting the `exdata3` PaLD graph, we want the layout to match the layout of the original data, so we will pass `as.matrix(exdata3)` to the `layout` parameter in the `plot_community_graphs()` function. Additionally, here the row names are meaningless, they just correspond to the location of the generated data, so we can remove the labels on the plot by setting `show_labels = FALSE`. The `only_strong` parameter will only display the strongly connected edges. Additionally, we can optionally pass parameters through the `...` to `plot::igraph` function; for example, to increase the vertex size we can set `vertex.size = 5`. Figure \ref{fig:fig6} shows this.

```{r fig6, fig.cap = "PaLD clustering of randomly generated example data (from Figure 4D from Berenhaut et al. (2022))"}
plot_community_graphs(exdata_cohesion_matrix,
                layout = as.matrix(exdata3),
                show_labels = FALSE, 
                only_strong = TRUE,
                vertex.size = 5)
```
